---
title: "Learning Video 1"
author: "Dean Stewart"
date: "16/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gapminder)
library(tidyverse)
library(readr)
library(tidyr)
library(dplyr)

data(gapminder)
gapminder_csv <- read_csv("../1_data/gapminder.csv")
gapminder_wide_csv <- read_csv("../1_data/gapminder_wide.csv")


```

## Making data tidy

unique variables for each column.
gapminder_wide is not like that...it has gdp_per_cap_1952, gdp_per_cap_1953...etc...

This is similar to what you often see on a website - with the data dates across the page.  eg. world bank or covid data for countries

Need to reshape...
```{r}
gap_long <- gapminder_wide_csv %>%
  gather(key, value, -continent, -country) %>%
  separate(key, c("variable_name","year"), "_") %>%
  spread(variable_name, value) %>%
  mutate(year=as.numeric(year))
  
# the first column is called 'key'.  the 'values' are called value.  Continent and country are ignored
# for more, see the video - this is explained
#spread variable_name, value - makes the variable names into columns, and then assigns the values to them.
```

Example using unite function
```{r}
gap_long2 <- gapminder_wide_csv %>%
  gather(key2, value, -continent, -country) %>%
  unite(key2, c(key2,country), sep="_") %>%
  separate(key2, c("variable_name","year", "country"), "_") %>%
  spread(variable_name, value) %>%
  mutate(year=as.numeric(year))
#this is a bit stupid - you unite the country into the key, and then separate it back out.  Just for demo of the unite function
```



## Assessing the data quality


```{r}
library(janitor)
library(psych)
library(dplyr)
library(gapminder)
data(gapminder)

#summarising the data

str(gapminder)  #str stands for structure
#tells dimensions, variable names, the data type of each variable, and sample values for each variable.

class(gapminder)
gapminder_adjnames <- gapminder
names(gapminder_adjnames) <- c("country","continent","year", "Life Exp", "population", "Gdp per Cap") #just change to 'not clean' header names
make.names(names(gapminder_adjnames)) #replaces spaces in names with full stop
clean_names(gapminder_adjnames)  # a better way - it replaces spaces with "_"

gapminder_clean2 <- rename(gapminder_adjnames, life_exp = "Life Exp")  #I only renamed one column to keep up with the video!
head(gapminder_clean2)


```
Table function
```{r}
table(gapminder$continent) #number of each continent
prop.table(table(gapminder$continent))  #tells the proportion of the total in each continent...ie count, but then divided by total

summary(gapminder) #returns counts for categorical, or boxplot stats if it's numeric
```

describe
```{r}
describe(gapminder) #psych package.  Also gives skewness and kurtosis.  Handy if we're using models that are normally distributed, for example.  if normal, you can just use linear regression, for example
```

Correlation
```{r}
cor(gapminder[,3:6])  #these are the numeric columns
cor(gapminder[,3:6], use="pairwise.complete.obs") #ignores na's

#default is pearson's correlation.  can also compute spearmans or kendles


```

counting rows and columns
use nrow() - helpful when splitting into training and testing dataset
and ncolumn()


