---
title: "manipulate cleanse and join"
author: "Dean Stewart"
date: "16/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Tibbles

Keeps the features of dataframes that are good and drops those that are bad!?

```{r}
my_tib <- tibble(x=1:3, y=c("blue","pink","yellow"))
my_tib
my_df = data.frame(x=1:3, y=c("blue","pink","yellow"))
my_df
as_tib = tibble(my_df)
as_tib

```
Unfortunately, this looks the same in a markdown!!
```{r}
data(starwars)
starwars #already a tibble
as.data.frame(starwars)
```

In a console, the tibble is much easier to work with.

Subsetting
```{r}
my_df[,1:2]
my_df[,1]   #no longer a data frame - now a vector
my_tib[,1:2]
my_tib[,1] #still a tibble
```

You can also progressively build in a tibble
```{r}
#anotherDF <- data.frame(nItems = c(12,45,107),
#                        cost = c(0.5,12,1.8),
#                        totalWorth = nItems * cost)  #produces an error because it can't find nItems

anothertib <- tibble(nItems = c(12,45,107),
                        cost = c(0.5,12,1.8),
                        totalWorth = nItems * cost)
anothertib
#this works
```

## now functions to clean and manipulate
```{r}
gapminder_all <- read_csv("../1_data/gapminderall.csv")
str(gapminder_all)
```

using keys to join
```{r}
gapminder_allwithkey <- gapminder_all %>% mutate(key = paste0(country, "_", year))
#country and year combined is helpful for joining later.
```

arrange
```{r}
gapminder_all_gdpcap <- gapminder_all %>%
  mutate(gdp_per_capita = gdp/population) %>%
  select(gdp_per_capita)# this works because it's a tibble
gapminder_all_gdpcap %>% arrange(gdp_per_capita) #like sort but does whole dataset
#can also do arrange(desc())
gapminder_all %>% arrange(fertility)
```

summarise
```{r}
gapminder_all %>% summarise(n_distinct(country))
length(unique(gapminder_all$country))

gapminder_all %>% filter(!is.na(population)) %>%
  summarise(mean_pop = mean(population), sd_pop = sd(population))

```
Pull or .$placeholder
```{r}
mean_pop <- gapminder_all %>% filter(!is.na(population)) %>%
  summarise(mean_pop = mean(population))
class(mean_pop)

#now, let's say we want to divide gdp in a country by the mean pop of all countries
gapminder_all2 <- gapminder_all %>%
  filter(!is.na(gdp))%>%
  mutate(gdpovermeanpop = gdp/mean_pop) %>%
  select(gdp,gdpovermeanpop)
  gapminder_all2 #if this isn't in markdown it shows a single value of 512.  problem is the class of mean_pop - it's a tibble or dataframe
  
#to convert to a value, we use pull or .$
  mean_pop <- gapminder_all %>%
    filter(!is.na(population)) %>%
    summarise(mean_pop = mean(population)) %>%
    pull(mean_pop)  #or .$mean_pop
    

```

```{r}
gapminder_all3 <- gapminder_all %>%
  filter(!is.na(gdp))%>%
  mutate(gdpovermeanpop = gdp/mean_pop) %>%
  select(gdp,gdpovermeanpop)
gapminder_all3  #much better now because mean_pop is a variable


```

group_by
```{r}
gapminder_all %>% group_by(continent) %>% filter(!is.na(gdp)) %>% summarise(average = mean(gdp))
#averages over the continent
```

distinct
```{r}
#more efficient than unique
distinct(gapminder_all)  #it's all distinct anyway..
distinct(gapminder_all, country, infant_mortality) #not unique

duplicated(gapminder_allwithkey$country)
table(gapminder_allwithkey$country[duplicated(gapminder_allwithkey$country)]) #each country duplicated 56 times...it occurs 57 times, duplicated 56 times.
```

duplicated for dataframe
```{r}
a<- gapminder_all[c(3,5),]
gapminder_dup <- rbind(gapminder_all,a)
gapminder_dup[duplicated(gapminder_dup),]  #watch that comma!  3rd and 5th obs is duplicated...of course

unique(gapminder_dup)
distinct(gapminder_dup) #does the same thing...
```

## Joins

```{r}
library(dplyr)
library(readr)
library(gapminder)
library(tibble)
data(gapminder)
gapminder_all <- read_csv("../1_data/gapminderall.csv")
```

joins
```{r}
gapminder_withkey <- gapminder %>% mutate(key = paste0(country, "_", year))
gapminder_allwithkey <- gapminder_all %>% mutate(key = paste0(country, "_", year))
dim(gapminder_withkey)
dim(gapminder_allwithkey)

#left join
gapminderleft <- left_join(gapminder_withkey, gapminder_allwithkey, by = "key")
dim(gapminderleft)

#right join
gapminderright <- right_join(gapminder_withkey, gapminder_allwithkey, by = "key")
dim(gapminderright) #the duplicated column names get .x or .y after them

#inner join
gapminderinner <- inner_join(gapminder_withkey, gapminder_allwithkey, by = "key")
dim(gapminderinner)

#full join (union)
gapminderfull <- full_join(gapminder_withkey, gapminder_allwithkey, by = "key")
dim(gapminderfull)

#semi join (inner join, but only with the x columns)
gapmindersemi <- semi_join(gapminder_withkey, gapminder_allwithkey, by = "key")
dim(gapmindersemi)


```

